Joe Biden Predicted to Win 56.59% of the Popular Vote in the 2024 U.S. Presidential Election

### 2024-03-17 paper/paper.qmd
final edits

The U.S. Presidential election will take place on Tuesday, November 5th, 2024, as America faces unprecedented levels of political polarization and diminished levels of trust in democratic institutions. As of March 11, 2024, former President Trump has 1078 out of the 1215 delegates needed to win the Republican nomination and President Biden has 1872 out of the 1968 delegates needed to win the Democratic nomination, therefore this election will feature the same candidates as the 2020 election. President Joe Biden will seek a second term and former President Donald Trump will try to become the second president to serve two non-consecutive terms. The only President to serve two non-consecutive terms was Grover Cleveland from 1885 to 1889 and 1893 to 1897 [@citeWaxman]. The 2024 election will define the future of women’s rights, LGBTQ+ rights, the separation of church and state, immigration policy, and foreign policy in America. America is at a turning point: Trump and Biden have very different ideologies, policy ideas, and visions for the future of the nation. Former President Trump has celebrated that he was “able to kill Roe v. Wade” [@citeKapur]. With 2022 Supreme Court decision to overturn Roe v. Wade, there will be nothing to stop Trump from signing a bill to restrict access to, or even criminalize, abortion, contraception, or IVF across the United States. In February 2024, Trump stated that he “would encourage” the Putin regime to do “whatever the hell they want” to NATO allies who don’t meet their defense spending targets (citation). On the contrary, President Biden has stood for women’s rights, protecting same-sex marriage, pro-immigration policies, and with Ukraine in their fight for freedom following Russia’s invasion. The January 6th, 2021, insurrection at the U.S. Capitol was a defining moment in the current era of political polarization in America. Trump refused to acknowledge his loss to Joe Biden in the 2020 U.S. presidential election. In August 2020, Trump claimed that the only way he was going to lose the election was “if the election is rigged.” [@citeAxelrod]. Since November 2020, he has falsely claimed that the election was fraudulent, and that Biden stole the election from him. On January 6, 2021, an angry pro-Trump mob stormed the U.S. Capitol in an attempt to "stop the steal," or in other words, to stop the joint session of congress from certifying Biden’s electoral college victory [@citeLonsdorf]. It is unclear whether former President Trump will concede if he is defeated again by President Biden in the 2024 election.

@@

The survey does not contain a question about a preferred 2024 Presidential Candidate. In fact, it was conducted in January 2024, before Super Tuesday took place and Donald Trump had secured almost enough delegates to win the Republican (GOP) nomination. As of March 11, 2024, former President Trump has secured 1078 out of the 1215 delegates needed to win the GOP nomination. Therefore, we were able to conclude that the two major candidates in the upcoming election will be incumbent President Joe Biden and former President Donald Trump. We created a `vote_biden` binary variable that indicates a respondent's preferred presidential candidate in the upcoming 2024 presidential election. It is equal to 1, which indicates that the respondent's preferred presidential candidate is Joe Biden, if their party affiliation is Democratic or if they voted for Joe Biden in 2020. `vote_biden` is equal to 0 otherwise, indicating that a respondent's preferred presidential candidate is Donald Trump. Donald Trump is arguably the most polarizing figure in American politics, therefore we are considering independents who voted for Joe Biden in 2020 to prefer Joe Biden over Donald Trump again in 2024. However, `vote_biden` will never be equal to 1 if the respondent voted for Donald Trump in 2016 or 2020. In the context of this report and the `vote_biden` variable, voting for a third-party candidate, a write-in vote, or not voting are not considered possible voting outcomes.

@@

Since the 1990s, America has developed an urban-rural ideological divide and it has kept growing [@citecornellians]. We created a new binary variable, `urban`, which is equal to "urban" if `urbancity2` indicates that the respondent lives in a big city, a smaller city, or a suburb, and "rural" if the respondent lives in a small town or rural area. This effectively distinguishes city and suburban voters from small town and rural voters, making our analysis simpler. We found that female survey respondents living in urban areas heavily favored Biden, while male survey respondents living in rural areas heavily favored Trump. In @fig-survey_urban_rural, we can see that while female survey respondents living in rural areas were split and male survey respondents living in urban areas favored Biden by a smaller margin, females in urban areas heavily favored Biden and males in rural areas heavily favored Trump.

@@

The sex, race, educ, and metro variables were cleaned and re-factored to mirror our survey data and to allow us to apply our model to our post-stratification data. @fig-demographics1 shows the demographic statistics of our survey and post-stratification data sets. The steps and code for producing @fig-demographics1 were obtained from @citeMatthew. We also created a `race_hispanic`, which is equal to either "hispanic" or "not hispanic" to ensure that our survey and post-stratification data sets would align, because this was a separate variable from race in our post-stratification data.

@@

@fig-datasetraces illustrates the demographic patterns in our post-stratification and survey data sets. The steps and code for producing @fig-stateprops were obtained from @citeMatthew. We can see that

@@

Overall, race demographics are consistent between our survey and post-stratification data sets. ,,, shows the exact count and proportion of respondents of different races in each data set. The proportion of Hispanic respondents in our survey and post-stratification data is comparable, at 12.7% and 13.5%, respectively. While there is a higher proportion of Hispanic respondents in our post-stratification data set, the difference is less than one percentage point. The proportions of white respondents in our survey (69.4%) and post-stratification (69.5%) data sets are nearly identical.

@fig-stateprops illustrates the distribution of voters across the 50 states, including Washington D.C. in both our survey and post-stratification data sets. The steps and code for producing @fig-stateprops were obtained from @citeMatthew.

@@

We can see that there are some slight differences in the proportions of respondents in the survey and post-stratification data from Alabama, California, Florida, Kansas, Maryland, Massachussets, Michigan, New York, Nebraska, and Virginia.

None of the voters in our survey data are from Vermont, but 0.198% of the voters in our post-stratification data are. Therefore, the results from applying our model to our post-stratification data will include a prediction for Vermont.

@fig-stateprops shows that there is a comparable distribution of voters across U.S. states in the survey and post-stratification data. However, the survey data over represents Florida, Michigan, New York, and Virginia.

@@

## Popular Vote Projection {#sec-first-point}

After performing MRP, we predicted that Joe Biden will wi

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

##  {#sec-second-point}

## Electoral College Projection

## Survey Error Hindering Model Accuracy {#sec-third-point}

## Weaknesses, Limitations, and Possible Next Steps {#sec-weaknesses-limitations}

The weaknesses of our model stem directly from the limited nature of the survey data set that we used, the America’s Political Pulse Survey Week 3 2024. As seen in @sec-survey, the electoral college predictions from Kansas (4 respondents), Wyoming (1 respondent), and Hawaii (7 respondents) are drastically different from the results seen in recent elections. There are very few respondents from smaller states with fewer electoral college votes, such as Kansas, Wyoming, Hawaii, New Hampshire, and Rhode Island in our survey data set. In addition to this, there are no respondents from Vermont in our survey data set. As mentioned, there were no respondents from Vermont and very few respondents from several states with few electoral college votes in our survey data set. As a result, we were not able to include state as a predictor in our model. Including state as a predictor would not have been applicable to respondents in our post-stratification census data from the state of Vermont at all.

@@

Logistic regression can only be used for binary classification. This limits the scope of this paper, as it only allows us to consider support for Joe Biden and Donald Trump in the upcoming 2024 presidential election. Our model does not consider support for third-party candidates, write-in votes, or the possibility of not voting, which could be a disadvantage when predicting results in swing states or for close elections. In the 2016 election, Donald Trump defeated Hillary Clinton by 11,612 votes (0.3 percentage points) in Michigan, which is considered a swing state. Third-party candidates Gary Johnson (Libertarian) and Jill Stein (Green) each received over 173,000 votes (3.6% of the popular vote) and 50,700 votes (1.1% of the popular vote) in Michigan [@citePolitico2016]. This is just one example of at least one third-party candidate receiving more votes than the difference in votes between the Democratic and Republican nominees, and it highlights a possible disadvantage of using logistic regression to predict vote preference.

An alternative option would be to fit a SoftMax regression model, which is a generalization of logistic regression that can predict the class probabilities for more than two classics and perform multi-class classification. However, a SoftMax regression model is more powerful than a logistic regression model. It would have overfit our survey data set because it only contains a total of 1000 observations, including less than five from some smaller states. While a SoftMax regression model might make sense for predicting vote preference when there are more than two candidates, this model would have been far too powerful for our survey data set and not worth the trade-offs. The bulk effective sample size and tail effective sample size would have been too low, which would mean that our posterior means, posterior medians, posterior variances, and tail quantiles would be unreliable [@citeStan]. We can justify using a logistic regression model because the United States has a two-party system, meaning that Joe Biden (Democrat) and Donald Trump (Republican) are the only candidates who can realistically win the presidency on November 5^th^. Although there will be third-party candidates, we do not yet know who they will be, nor are they expected to win any electoral college votes.

With a survey data set that is as limited as the one we currently have, SoftMax regression is not a viable option. However, the possibility of third-party candidates receiving vote shares larger than the margin of victory of the Democratic or Republican nominee in key battleground states highlights the need to consider the fact that a small percentage of voters will cast a ballot for someone not named Donald Trump or Joe Biden on November 5th. The first step towards extending and improving this report in the future would be to find a survey data set that has significantly more respondents, is somewhat reflective of voting trends and demographics in each state, and specifically asks respondents who they plan to vote for in the 2024 U.S. presidential election. Once we address the issue of having very few, or even no respondents from smaller states and no question about 2024 vote intention, we will be able to build a stronger and more accurate model to predict vote choice based on state, urban or rural area, demographics such as sex, age, and race, and education. A robust survey data set with enough respondents (possibly upwards of 10,000) will allow us to train a more powerful SoftMax regression model to perform multi-class classification and classify voters as Trump, Biden, third-party, or abstaining voters using similar predictors to the ones outlined in @sec-model as well as state.

@@

## Additional Data Details

## Data Cleaning

## Additional Tables

@tbl-racetables

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| include: true
#| label: tbl-racetables
#| tbl-cap: Races of survey and post stratification data respondents
#| tbl-subcap: ["Survey data", "Post-stratification data"]
#| layout-ncol: 2

races_table <- survey_analysis_data |>
  group_by(races) |>
   summarise(Count = n()) |>
  mutate("%" = round(Count / nrow(survey_analysis_data) * 100, 2))

races_table <- races_table |> rename(Race = races)

kable(races_table)

races2 <- poststrat_analysis_data |>
  group_by(races) |>
  summarise(Count = n())|>
  mutate("%" = round(Count / nrow(poststrat_analysis_data) * 100, 2))

races2 <- races2 |> rename(Race = races)

kable(races2)
```

## Additional Data Details

@@

@fig-stanareyouokay-2 shows the trace plot for our intercept. We can see that the lines bounce, but remain horizontal and have a lot of overlap with each other [@tswd]. @fig-tracesex, @fig-traceage, @fig-tracerace, @fig-traceeducation, and @fig-traceurban show the trace plots for our sex, age_bracket, races, education_level, and urban variables. Each of these trace plots looks similar to @fig-stanareyouokay-2, as the lines bounce, remain horizontal, and overlap a lot. Our trace plots suggest that the distribution behaves as expected, and we do not need to re-run our model with fewer predictors or different priors [@tswd].

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Rhat plot", "Trace plot for intercept"]
#| layout-ncol: 1

@@

plot(presidential_vote_model, "trace","(Intercept)")

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-tracesex
#| fig-cap: "Trace plots for sex"

plot(presidential_vote_model, "trace", "sexmale")

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-traceage
#| fig-cap: "Trace plot for age_bracket"
#| fig-subcap: ["Trace plot for age_bracket30-44", "Trace plot for age_bracket45-59", "Trace plot for age_bracket60+"]
#| layout-ncol: 3

plot(presidential_vote_model, "trace", "age_bracket30-44") 
plot(presidential_vote_model, "trace", "age_bracket45-59") 
plot(presidential_vote_model, "trace", "age_bracket60+") 
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-tracerace
#| fig-cap: "Trace plot for races"
#| fig-subcap: ["Trace plot for racesblack", "Trace plot for racesmixed", "Trace plot for racesother", "Trace plot for racesnative american", "Trace plot for raceswhite", "Trace plot for race_hispanic not hispanic"]
#| layout-ncol: 3

par(mfrow=c(2, 3))

plot(presidential_vote_model, "trace", "racesblack") 
plot(presidential_vote_model, "trace", "racesmixed") 
plot(presidential_vote_model, "trace", "racesnative american") 
plot(presidential_vote_model, "trace", "racesother") 
plot(presidential_vote_model, "trace", "raceswhite")
plot(presidential_vote_model, "trace", "race_hispanicnot hispanic")

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-traceeducation
#| fig-cap: "Trace plot for education_level"
#| fig-subcap: ["Trace plot for education_levelNo HS", "Trace plot for education_levelHigh school graduate", "Trace plot for education_levelSome college", "Trace plot for education_level4-year", "Trace plot for education_levelPost-grad"]
#| layout-ncol: 1

plot(presidential_vote_model, "trace", "education_levelNo HS") 
plot(presidential_vote_model, "trace", "education_levelHigh school graduate") 
plot(presidential_vote_model, "trace", "education_levelSome college") 
plot(presidential_vote_model, "trace", "education_level4-year") 
plot(presidential_vote_model, "trace", "education_levelPost-grad") 
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-traceurban
#| fig-cap: "Trace plot for urban"

plot(presidential_vote_model, "trace", "urbanurban") 
```

@@

In @fig-modelplot, we show the 90% credibility intervals for the predictors of `vote_biden`. We see that the 90% credibility interval for when urban=urban is greater than 0. This means that a respondent who lives in an urban area is more likely to support Joe Biden. We also see that the 90% credibility interval for sex=male is below zero. This means that based on our survey data and model, male respondents are more likely to support Donald Trump.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-modelplot
#| fig-cap: "Credibility intervals for predictors of vote_biden"
modelplot(presidential_vote_model, conf_level = 0.9) +
  labs(x = "90% credibility interval")
```

>>>>

The U.S. Presidential election will take place on Tuesday, November 5th, 2024, as America faces unprecedented levels of political polarization and diminished levels of trust in democratic institutions. As of March 11, 2024, former President Trump has 1078 out of the 1215 delegates needed to win the Republican nomination and President Biden has 1872 out of the 1968 delegates needed to win the Democratic nomination, therefore this election will feature the same candidates as the 2020 election. President Joe Biden will seek a second term and former President Donald Trump will try to become the second president to serve two non-consecutive terms. The only President to serve two non-consecutive terms was Grover Cleveland from 1885 to 1889 and 1893 to 1897 [@citeWaxman]. The 2024 election will define the future of women’s rights, LGBTQ+ rights, the separation of church and state, immigration policy, and foreign policy in America. America is at a turning point: Trump and Biden have very different ideologies, policy ideas, and visions for the future of the nation. Former President Trump has celebrated that he was “able to kill Roe v. Wade” [@citeKapur]. With 2022 Supreme Court decision to overturn Roe v. Wade, there will be nothing to stop Trump from signing a bill to restrict access to, or even criminalize, abortion, contraception, or IVF across the United States. In February 2024, Trump stated that he “would encourage” the Putin regime to do “whatever the hell they want” to NATO allies who don’t meet their defense spending targets [@citeSullivan]. On the contrary, President Biden has stood for women’s rights, protecting same-sex marriage, pro-immigration policies, and with Ukraine in their fight for freedom following Russia’s invasion. The January 6th, 2021, insurrection at the U.S. Capitol was a defining moment in the current era of political polarization in America. Trump refused to acknowledge his loss to Joe Biden in the 2020 U.S. presidential election. In August 2020, Trump claimed that the only way he was going to lose the election was “if the election is rigged” [@citeAxelrod]. Since November 2020, he has falsely claimed that the election was fraudulent, and that Biden stole the election from him. On January 6, 2021, an angry pro-Trump mob stormed the U.S. Capitol in an attempt to "stop the steal," or in other words, to stop the joint session of congress from certifying Biden’s electoral college victory [@citeLonsdorf]. It is unclear whether former President Trump will concede if he is defeated again by President Biden in the 2024 election.

@@

The survey does not contain a question about a preferred 2024 Presidential Candidate. In fact, it was conducted in January 2024, before Super Tuesday took place and Donald Trump had secured almost enough delegates to win the Republican (GOP) nomination. We created a variable `vote_biden`, which is equal to 1 if a respondent's preferred 2024 presidential candidate is Joe Biden, and 0 if it is Donald Trump. The data cleaning process used for the creation of the `vote_biden` variable is outlined in @sec-datacleaning.

@@

Since the 1990s, America has developed an urban-rural ideological divide and it has kept growing [@citecornellians]. We created a new binary variable, `urban`, indicates whether the respondent lives in an urban or rural area. The data cleaning steps that we used to create this variable are outlined in @sec-datacleaning. First, we added the `urban` variable to our survey data. It is equal to "urban" if `urbancity2` indicates that the respondent lives in a big city, a smaller city, or a suburb, and "rural" if the respondent lives in a small town or rural area. This effectively distinguishes city and suburban voters from small town and rural voters, making our analysis simpler. We found that female survey respondents living in urban areas heavily favored Biden, while male survey respondents living in rural areas heavily favored Trump. In @fig-survey_urban_rural, we can see that while female survey respondents living in rural areas were split and male survey respondents living in urban areas favored Biden by a smaller margin, females in urban areas heavily favored Biden and males in rural areas heavily favored Trump.

@@

The sex, race, educ, and metro variables were cleaned and re-factored to mirror our survey data and to allow us to apply our model to our post-stratification data. @fig-demographics1 shows the demographic statistics of our survey and post-stratification data sets. The steps and code for producing @fig-demographics1 were obtained from @citeMatthew. We also created a `race_hispanic`, which is equal to either "hispanic" or "not hispanic" to ensure that our survey and post-stratification data sets would align, because this was a separate variable from race in our post-stratification data. The data cleaning steps used to create the `urban` variable to make our post-stratification data set mirror our survey data set are outlined in @sec-datacleaning.

@@

@fig-datasetraces illustrates the demographic patterns in our post-stratification and survey data sets. The steps and code for producing @fig-stateprops were obtained from @citeMatthew.

@@

Overall, race demographics are consistent between our survey and post-stratification data sets. The proportion of Hispanic respondents in our survey and post-stratification data is comparable, at 12.7% and 13.5%, respectively. While there is a higher proportion of Hispanic respondents in our post-stratification data set, the difference is less than one percentage point. The proportions of white respondents in our survey (69.4%) and post-stratification (69.5%) data sets are nearly identical. @fig-stateprops illustrates the distribution of voters across the 50 states, including Washington D.C. in both our survey and post-stratification data sets. The steps and code for producing @fig-stateprops were obtained from @citeMatthew.

@@

We can see that there are some slight differences in the proportions of respondents in the survey and post-stratification data from Alabama, California, Florida, Kansas, Maryland, Massachussets, Michigan, New York, Nebraska, and Virginia. None of the voters in our survey data are from Vermont, but 0.198% of the voters in our post-stratification data are. Therefore, the results from applying our model to our post-stratification data will include a prediction for Vermont. @fig-stateprops shows that there is a comparable distribution of voters across U.S. states in the survey and post-stratification data. However, the survey data over represents Florida, Michigan, New York, and Virginia.

@@

## How we produced our popular vote and electoral college predictions {#sec-first-point}

## The popular vote does not necessarily define the electoral college

When discussing the results of our model, it is crucial to consider the quality of the survey data that was used to fit it. Despite projecting Joe Biden to win an impressing, but not dominant, 56.59% of the popular vote in the 2024 presidential election when applied to the post-stratification data, our model produced a mean estimate of 471 electoral college votes for Joe Biden. We know that winning a certain percentage of the popular vote does not translate into a fixed numer of electoral college votes. Instead, the candidate with the most votes in each state receives all of its electoral college votes, and in America's two-party system, the candidate whose electoral votes add up to 270 or more wins the presidency. So, in theory, it is possible to win 538 electoral votes with exactly 50.1% of the vote in each state. However, this scenario is extremely unlikely. It is also possible to win the popular vote, but still lose the electoral college and the presidency. In fact, Hillary Clinton won the popular vote but lost the electoral college to Donald Trump, 232 to 306, in 2016.

The shortcomings of our survey data set are highlighted in @fig-estimatedpropsupport2, where we can see that there is no visible pattern or relationship between predictions for the percent of voters that will support Biden in each state based on the survey and post-stratification data sets. The gray dots, which represent our prediction for the percent of voters who prefer Biden in each state based on our survey data set, are scattered all over the place. We used that same survey data set to fit the model that was applied to our post-stratification data set to produce the blue (where the mean prediction for Biden support was over 50%) and red (where the mean prediction for Biden support was below 50%) prediction intervals that show which candidate would win each state's electoral college votes. While state was not a predictor for our model, it is clear that in many cases, our survey and post-stratification data sets produced very different predictions for the percentage of votes that Biden will receive in many states.

The lower estimate for the number of electoral college votes that Biden will receive based on the results of our MRP in November is 363, while the upper estimate is 538, as shown in @tbl-popvotepost-popularvote. That is a difference of 175 seats, which gives us little confidence in our current model's ability to accurately predict the electoral college. In comparison, our prediction for the number of electoral college votes that President Biden will win in the 2024 election based on our survey data is 324, as shown in @tbl-surveyresults. 471 electoral college votes is not even relatively comparable to the 306 that he received in 2020, or the number of electoral college votes that presidents have received in recent elections [@citeCNN]. Therefore, there is reason for us to believe that the seemingly random and far-off predictions for the proportion of votes that Biden will receive in numerous states based on our survey data limits our model’s ability to accurately predict the electoral college.

## Weaknesses and Limitations {#sec-weaknesses-limitations}

Our model can only ever be as good as the data set that we used to build and fit it [@tswd]. The weaknesses of our model stem directly from the limited nature of the survey data set that we used, the America’s Political Pulse Survey Week 3 2024. As seen in @sec-survey, the electoral college predictions from Kansas (4 respondents), Wyoming (1 respondent), and Hawaii (7 respondents) are drastically different from the results seen in recent elections. There are very few respondents from smaller states with fewer electoral college votes, such as Kansas, Wyoming, Hawaii, New Hampshire, and Rhode Island in our survey data set. In addition to this, there are no respondents from Vermont in our survey data set. As mentioned, there were no respondents from Vermont and very few respondents from several states with few electoral college votes in our survey data set. As a result, we were not able to include state as a predictor in our model. Including state as a predictor would not have been applicable to respondents in our post-stratification census data from the state of Vermont at all.

@@

Logistic regression can only be used for binary classification. This limits the scope of this paper, as it only allows us to consider support for Joe Biden and Donald Trump in the upcoming 2024 presidential election. Our model does not consider support for third-party candidates, write-in votes, or the possibility of not voting, which could be a disadvantage when predicting results in swing states or for close elections. In the 2016 election, Donald Trump defeated Hillary Clinton by 11,612 votes (0.3 percentage points) in Michigan, which is considered a swing state. Third-party candidates Gary Johnson (Libertarian) and Jill Stein (Green) each received over 173,000 votes (3.6% of the popular vote) and 50,700 votes (1.1% of the popular vote) in Michigan [@citePolitico2016]. This is just one example of at least one third-party candidate receiving more votes than the difference in votes between the Democratic and Republican nominees in a key battleground state. It highlights a major limitation of using logistic regression to predict vote preference.

## Next Steps

We can address our model's inability to consider more than two voting outcomes by fitting a SoftMax regression model. SoftMax regression is a generalization of logistic regression that can predict the class probabilities for more than two classics and perform multi-class classification. However, a SoftMax regression model is more powerful than a logistic regression model. It would severely overfit our current survey data set because it only has 1000 observations, including less than five from some smaller states. While a SoftMax regression model might make sense for predicting vote preference when there are more than two candidates, this model would have been far too powerful for our survey data set and not worth the trade-offs. The bulk effective sample size and tail effective sample size would have been too low, which would mean that our posterior means, posterior medians, posterior variances, and tail quantiles would be unreliable [@citeStan]. We can justify using a logistic regression model because the United States has a two-party system, meaning that Joe Biden (Democrat) and Donald Trump (Republican) are the only candidates who can realistically win the presidency on November 5^th^. Although there will be third-party candidates, we do not yet know who they will be, nor are they expected to win any electoral college votes.

With a survey data set that is as limited as the one we currently have, SoftMax regression is not a viable option. However, the possibility of third-party candidates receiving vote shares larger than the margin of victory of the Democratic or Republican nominee in key battleground states highlights the need to consider the fact that a small percentage of voters will cast a ballot for someone not named Donald Trump or Joe Biden on November 5th. Therefore, the first step towards extending and improving this report in the future would be to find a survey data set that has significantly more respondents, is somewhat reflective of voting trends and demographics in each state, and specifically asks respondents who they plan to vote for in the 2024 U.S. presidential election. Once we address the issue of having very few, or even no respondents from smaller states and no question about 2024 vote intention, we will be able to build a stronger and more accurate model to predict vote choice based on state, urban or rural area, demographics such as sex, age, and race, and education. A robust survey data set with enough respondents (possibly upwards of 10,000) will allow us to train a more powerful SoftMax regression model to perform multi-class classification and classify voters as Trump, Biden, third-party, or abstaining voters using similar predictors to the ones outlined in @sec-model as well as state. We believe that we can extend and improve our study by using SoftMax regression, but it is crucial for us to find a better survey data set first.

@@

## Data Cleaning {#sec-datacleaning}

We created a binary variable, `vote_biden` to indicate whether each survey respondent's preferred 2024 presidential candidate is Joe Biden or Donald Trump. `vote_biden` is equal to 1, which indicates that the respondent's preferred presidential candidate is Joe Biden, if their party affiliation is Democratic or if they voted for Joe Biden in 2020. `vote_biden` is equal to 0 otherwise, indicating that a respondent's preferred presidential candidate is Donald Trump. Donald Trump is arguably the most polarizing figure in American politics, therefore we are considering independents who voted for Joe Biden in 2020 to prefer Joe Biden over Donald Trump again in 2024. However, vote_biden will never be equal to 1 if the respondent voted for Donald Trump in 2016 or 2020. In the context of this report and the vote_biden variable, voting for a third-party candidate, a write-in vote, or not voting are not considered possible voting outcomes.

We also created a new variable, `urban`. First, we added the `urban` variable to our survey data. It is equal to "urban" if `urbancity2` indicates that the respondent lives in a big city, a smaller city, or a suburb, and "rural" if the respondent lives in a small town or rural area. For our post-stratification data set, we also gave the `urban` variable two levels, "urban" and "rural" to align with our survey data set. If the `metro` variable indicated that a census respondent lives in a metropolitan area, whether inside or outside of the principal/central city, `urban` is equal to "urban," and we set it to rural for census respondents who do not live in a metropolitan area.

@@

@fig-stanareyouokay-2 shows the trace plots for our predictors. We can see that the lines bounce, but remain horizontal and have a lot of overlap with each other [@tswd]. This suggest that the distribution behaves as expected, and we do not need to re-run our model with fewer predictors or different priors [@tswd].

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Rhat plot", "Trace plot"]
#| layout-ncol: 1

@@

plot(presidential_vote_model, "trace")

@@

@@

In @fig-modelplot, we show the 90% credibility intervals for the predictors of `vote_biden`.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-modelplot
#| fig-cap: "90% Credibility intervals for predictors of vote_biden"
modelplot(presidential_vote_model, conf_level = 0.9) +
  labs(x = "90% credibility interval")
```
